---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

#Reads the data

```{r}
data = read.csv('activity.csv')
data$date = as.Date(data$date, "%Y-%m-%d")
splitDay = split(data, data$date)
```

## What is mean total number of steps taken per day?

```{r}
totalSteps = sapply(splitDay, function(x) sum(x$steps))
medianSteps = sapply(splitDay, function(x) median(x$steps))
hist(totalSteps,main="Histogram of Frequency of TotalSteps/Day")
```

# Total numbers of steps/day

```{r}
totalSteps
```

# Median number of steps/day

```{r}
medianSteps
```


## What is the average daily activity pattern?

```{r}
interval = seq(0,2355,5)
steps = numeric()
for(i in 1:472) {
        steps[i] = 0
}
avgSteps = data.frame(interval, steps)
names(avgSteps) = c('interval','avgsteps')
    
    
for (i in 1:472){
        avgSteps$avgsteps[i] = mean(data$steps[data$interval == interval[i]], na.rm=TRUE)
}
    
   
plot(avgSteps, type='l', main='Avg. Steps over 5 minute intervals',xlab='Interval(minutes)', ylab='Average Steps')
    
    
cat('The interval with the highest average is' , avgSteps$interval[which.max(avgSteps$avgsteps)])
```


## Imputing missing values

```{r}
numNa = sum(is.na(data$steps))
cat('The number of NA values in the dataset is' , numNa)
```

# All NA values will be replaced with 0

```{r}
noNa = data
noNa[is.na(noNa)]=0
splitNa = split(noNa, noNa$date)
totalNa = sapply(splitNa, function(x) sum(x$steps))
hist(totalNa,main="Histogram of Frequency of TotalSteps/Day", xlab="Total number of Steps")
avgNa = sapply(splitNa, function(x) mean(x$steps))
medNa = sapply(splitNa, function(x) median(x$steps))
```

# The following is the adjusted average values after changing all NA values to 0

```{r}
avgNa
```

# The following is the adjusted median values after chaning all NA values to 0

```{r}
medNa
```

## Are there differences in activity patterns between weekdays and weekends?

# The following creates subsets containing the weekday/weekend values of the dataset

```{r}
weekData = data
    weekData[is.na(weekData)]=0    
    weekData$days = weekdays(weekData$date)
    weekData$days = as.factor(ifelse(weekdays(weekData$date) %in% c("Saturday","Sunday"), 
                                      "Weekend", "Weekday")) 

    avgEnd = mean(weekData$steps[weekData$days == 'Weekend'])
    avgDay = mean(weekData$steps[weekData$days == 'Weekday'])

    
    weekdays = subset(weekData, weekData$days == 'Weekday')
    weekends = subset(weekData, weekData$days == 'Weekend')
    
    weekend = numeric()
    weekday = numeric()
    for(i in 1:472) {
        weekday[i] = 0
        weekend[i] = 0
    }
    avgWeek = data.frame(interval, weekend, weekday)
    names(avgWeek) = c('interval','avgsteps.weekend','avgsteps.weekday')
    
    for (i in 1:472){
        avgWeek$avgsteps.weekday[i] = mean(weekdays$steps[weekdays$interval == interval[i]], na.rm=TRUE)
        avgWeek$avgsteps.weekend[i] = mean(weekends$steps[weekends$interval == interval[i]], na.rm=TRUE)
    }
```

# The plot is setup to show the differences in the average steps vs time period between weekends and weekdays

```{r}
par(mfrow=c(2,1))
    plot(avgWeek$interval, avgWeek$avgsteps.weekend, type='l', main ="Weekend vs Avg Steps",
         xlab="Interval(steps)", ylab="Average Steps")
    plot(avgWeek$interval, avgWeek$avgsteps.weekday, main ="Weekday vs Avg Steps", 
         xlab="Interval(steps)", ylab="Average Steps", type='l')

```
